Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-04 paper.tex
fixing compile, rightarrow needs to be in math mode
#7 ok done with this read, i like it, several comments anyway of course

On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting which of the currently \emph{occupant} subgroups on a multiprocessor for execution. The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources. Effective virtualization requires \emph{oversubcription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available. As a result, a single multiprocessor may host multiple workgroups simultaneously. However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch.

@@

\subsubsection{Fairness and Progress}
Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among subgroups---and \emph{progress guarantees}---ensuring that subgroups eventually make progress towards termination. However, the term "fairness" is used differently across the literature, leading to potential confusion. Sorensen et al.\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees. In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups. In this work, we adopt NVIDIA's terminology.

\thomas{A scheduler may be fair but may not have FPG (M1 Pro for example); a scheduler may have FPG but may not be fair (ARM Mali). \newline
  Sorensen calls FPG "fairness:" "We have described two schedulers: fair and unfair, under which starvation-freedom for blocking idioms is either always or never guaranteed."\newline
  Nvidia, by contrast uses fairness in line with our language. From Tesla architecture: \newline
  "A scoreboard qualifies each warp for issue each cycle. The instruction scheduler prioritizes all ready warps and selects the one with highest priority for issue. Prioritization considers warp type, instruction type, and 'fairness' to all warps executing in the SM\@." \newline
  From DecoupledLookback: \newline
  "Safety. The algorithm will run to completion if the system guarantees forward-progress for all processors." \newline
  "Blocking will be minimal for systems that provide fair or nearly-fair scheduling. Fairness ensures that all processors will have recorded their aggregates in roughly the same amount of time."
}

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). Although \( \oplus \) must be associative, it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

@@

Thus, oversubcription and \emph{Chained-Scan} emerges as the preferred architecture, as it naturally aligns with portability by leveraging hardware schedulers to dynamically manage occupancy across diverse architectures without requiring developer intervention. Furthermore, as a single kernel with inherently serial ordering, \emph{Chained-Scan} is suitable for \emph{in-situ} operations, augmenting practical routines such as data deduplication, sparse data compaction, error correction, and interval merging.

@@

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (from possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant. However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE\@.}

@@

Although the serial assignment of work tiles ensures that predecessor tiles are scheduled or completed, it does not guarantee their reductions will be available when a workgroup begins \emph{Chained Propagation}. This uncertainty arises because workgroups are scheduled \emph{individually}, meaning a predecessor may not have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup spins, consuming execution resources while waiting. Because workgroups may share execution resources by being co-resident on the same multiprocessor, without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing progress and creating a circular dependency, satisfying condition 4b and resulting in deadlock.

For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor. Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG@. Both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

@@

\subsection{Earlier Attempts at Portability}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{Decoupled-Lookback} approach without FPG requirements in \emph{Scalar Fallback}. Each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. This eliminates the \emph{mutual exclusion} condition necessary for deadlock, as workgroups can operate on data from other tiles. While this may introduce redundant work when a tile is late rather than truly blocking, it guarantees termination by allowing any workgroup to complete a blocking tile’s reduction, breaking inter-workgroup dependencies.

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup enters a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Additionally, since only one thread loads elements, the associated cache line is severely underutilized. Second, once the fallback completes, the workgroup does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.

@@

To optimize the fallback operation, we employ three strategies. Although \emph{Chained-Propagation} is performed by a single thread or subgroup, we want the entire workgroup to participate in the fallback. First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary. Second, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback. While this approach risks redundant fallbacks---since a late tile is indistinguishable from a genuinely blocking one---exposing the maximum spin count as a tunable parameter helps minimize false positives. Finally, after completing a fallback, the workgroup attempts to insert the reduction in place of the blocking tile, preventing redundant fallbacks by subsequent workgroups.

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$, independent of FPG, due to \emph{false-positive} fallback operations. As a result, memory bandwidth usage increases (and throughput decreases) from $O(fn)$ to $O((2 + f)n)$. However, as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. As we later empirically demonstrate, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

\section{Implementation}
At a high level, our algorithmn follows a five phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking subgroup scans,
  allowing the subgroup-block scan to remain barrier-free and in registers.The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the local spine to $\frac{w}{s}$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation, engaging in fallback operations if necessary. Once completed, it posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase is negligible due to our overall thriftiness, including the deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

\subsection{Decoupled-Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile comprises a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently, each responsible for updating its own tile state but may opportunistically update preceding tiles in the event of a fallback. Tile states transition strictly forward---Not Ready \rightarrow Ready \rightarrow Inclusive---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\epsilon$. State transitions dictate reduction accumulation: at \emph{Ready}, local reductions are posted; at \emph{Inclusive}, preceding values are incorporated. To coordinate this process, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

@@

  \item \textbf{Initialization:}
        \begin{itemize}
          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained-Propagation} phase.
          \item The first thread initializes a \emph{previous reduction} variable in registers with $\epsilon$ to hold the accumulated reductions of preceding tiles.
          \item To traverse the spine, the first thread initializes a \emph{current tile} pointer to its immediate predecessor.  
        \end{itemize}

  \item \textbf{Lookback:}
        \begin{itemize}
          \item The first thread queries the state of \emph{current tile}.
                \begin{itemize}
                  \item If the tile is in the \emph{Ready} state, the workgroup ingests the value and sets \emph{current tile} to the preceding tile, repeating this step until reaching a tile that is \emph{Not Ready}.
                  \item If the tile is in the \emph{Inclusive} state, the workgroup ingests the value and proceeds directly to step 4.
                  \item If the tile is in the \emph{Not Ready} state, the first thread spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}. 
                        \begin{itemize}
                            \item If this count is exceeded, it exits the lookback loop, leaving the control flag \emph{Locked} to signal to the workgroup that a fallback is required. Additionally, it copies \emph{current tile} into shared memory to broadcast the fallback location to the workgroup.
                            \item Else, it continues spinning until the tile transitions to a \emph{Ready} or \emph{Inclusive} state. Without FPG, \emph{Decoupled Lookback} risks spinning indefinitely and deadlocking at this stage.
                        \end{itemize}
                \end{itemize}
        \end{itemize}

>>>>

On GPU hardware, scheduling is divided into two levels: a \emph{workgroup} scheduler, which manages kernel launches and maps virtual processors, workgroups, to physical processors, \emph{multiprocessors}, and a \emph{subgroup} scheduler, responsible for selecting which of the currently \emph{occupant} subgroups on a multiprocessor for execution. The GPU programming model virtualizes processors to ensure GPU programs---\emph{kernels}---remain portable across hardware with differing physical resources. Effective virtualization requires \emph{oversubscription}, a programming paradigm where tasks are (over)partitioned in such a way that kernels request more virtual processors than are physically available. As a result, a single multiprocessor may host multiple workgroups simultaneously. However, as a consequence of virtualization, the developer relinquishes control over the order in which workgroups are launched, and once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch.

@@

\subsubsection{Fairness and Progress}
Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among subgroups---and \emph{progress guarantees}---ensuring that subgroups eventually make progress towards termination. However, the term "fairness" is used differently across the literature, leading to potential confusion. Sorensen et al.~\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees. In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups. In this work, we adopt NVIDIA's terminology.

\thomas{A scheduler may be fair but may not have FPG (M1 Pro for example); a scheduler may have FPG but may not be fair (ARM Mali). \newline
  Sorensen calls FPG "fairness:" "We have described two schedulers: fair and unfair, under which starvation-freedom for blocking idioms is either always or never guaranteed."\newline
  NVIDIA, by contrast uses fairness in line with our language. From Tesla architecture: \newline
  "A scoreboard qualifies each warp for issue each cycle. The instruction scheduler prioritizes all ready warps and selects the one with highest priority for issue. Prioritization considers warp type, instruction type, and 'fairness' to all warps executing in the SM\@." \newline
  From DecoupledLookback: \newline
  "Safety. The algorithm will run to completion if the system guarantees forward-progress for all processors." \newline
  "Blocking will be minimal for systems that provide fair or nearly-fair scheduling. Fairness ensures that all processors will have recorded their aggregates in roughly the same amount of time."
}

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

@@

Thus, oversubscription and \emph{Chained-Scan} emerges as the preferred architecture, as it naturally aligns with portability by leveraging hardware schedulers to dynamically manage occupancy across diverse architectures without requiring developer intervention. Furthermore, as a single kernel with inherently serial ordering, \emph{Chained-Scan} is suitable for \emph{in-situ} operations, augmenting practical routines such as data deduplication, sparse data compaction, error correction, and interval merging.

@@

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (from possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers, classifying \emph{Chained-Scan} as FPG-reliant. However, their work primarily focuses on formalizing progress models rather than their direct application to specific algorithms.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline ``Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated'' \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE\@.} \john{I don't have an opinion on this, you're familiar with this and I'm not, but this seems like a good footnote. Or we can ask Tyler.}

@@

Although the serial assignment of work tiles ensures that predecessor tiles are scheduled or completed, it does not guarantee their reductions will be available when a workgroup begins \emph{Chained Propagation}. This uncertainty arises because workgroups are scheduled \emph{individually}, meaning a predecessor may not have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup spins, consuming execution resources while waiting. Because workgroups may share execution resources by being co-resident on the same multiprocessor\john{Question, and maybe this is worth a footnote. What if we guaranteed that no multiprocessor ever had more than one workgroup scheduled to it, say by always allocating the maximum amount of workgroup memory? Of course there would be performance issues. But does that solve the problem? Or do we descend into an analogous issue with subgroup scheduling?}, without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing progress and creating a circular dependency, satisfying condition 4b and resulting in deadlock.

For example, consider a scan operation requiring two workgroups on a GPU with a single multiprocessor. Both workgroups are co-resident, but the multiprocessor has only one scheduling unit, which lacks FPG\@. Both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, repeatedly querying the memory address for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it continues scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

@@

\subsection{Earlier Attempts at Portability}
\john{I'd like Raph to chime in here. What I worry about in the way this subsection is presented is that a reviewer will say ``Oh, the fallback technique was already invented in 2021, all this paper does is makes it a little better''. On one hand I totally want Raph to get credit for this cool idea. On the other hand, writing this subsection in this way, as a citable previous publication, significantly reduces the size of the advance that this paper we're writing hopes to capture. If I'm being selfish and trying to make this paper look as good as possible, I downplay the previous work a little and don't present it as if it was a previous peer-reviewed publication. At the least I want to use the words ``blog post'' in the text here.}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{Decoupled-Lookback} approach without FPG requirements in \emph{Scalar Fallback}. Each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. This eliminates the \emph{mutual exclusion} condition necessary for deadlock, as workgroups can operate on data from other tiles. While this may introduce redundant work when a tile is late rather than truly blocking, it guarantees termination by allowing any workgroup to complete a blocking tile’s reduction, breaking inter-workgroup dependencies.

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup enters a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Additionally, since only one thread loads elements, the associated cache line is severely underutilized. Second, once the fallback completes, the workgroup does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. \john{If we can, we would like to [politely] say that this previous work was missing a key insight, e.g., while fallback can address the portability problem, \ldots\ Would we say that it's not decoupled in the way that the technique we present in this paper is? Or is the only deficiency the choice of a serial fallback?} While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.

@@

To optimize the fallback operation, we employ three strategies. Although \emph{Chained-Propagation} is performed by a single thread or subgroup, for performance reasons we want the entire workgroup to participate in the fallback. (1) First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary. (2) Next, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback. While this approach risks redundant fallbacks---since a late tile is indistinguishable from a genuinely blocking one---exposing the maximum spin count as a tunable parameter helps minimize false positives. \john{I can see it is sensible to have a maximum spin count but I do not see how the action of exposing the maximum spin count to the programmer minimizes false positives.} (3) Finally, after completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.

The viability of \emph{Decoupled-Fallback} relies on the same cache locality properties as \emph{Decoupled-Lookback}. Since no subgroup scheduler is perfectly fair, all hardware inherently has a natural fallback rate $f$\!, independent of FPG, due to \emph{false-positive} fallback operations. As a result, memory bandwidth usage increases (and throughput decreases) from $O(fn)$ to $O((2 + f)n)$. However, as serial ordering confines spine memory accesses within an $o$-width sliding window, scan input accesses are likewise constrained to an $ot$-width window, increasing their likelihood of residing in cache. As we later empirically demonstrate, this cache residency significantly reduces memory access costs, making fallback operations predominantly compute-bound when reducing a blocking tile. Since pure reduction has minimal overhead, throughput approaches $O(2n)$ on most devices.

\section{Implementation}
At a high level, our algorithm follows a five-phase construction:
\begin{enumerate}
  \item[(0)] \textbf {Initialization}: The first thread in the workgroup atomically acquires its work-tile by atomically incrementing a value in global memory.
  \item \textbf{Load Phase}: Each thread loads $p$ vectors, consuming the input in a subgroup-blocked ordering. For a subgroup size $s$ and vector size $v$, each subgroup processes a contiguous block of size $s \cdot p \cdot v$.
  \item \textbf{Subgroup-Rake}: Each thread scans its vectors privately, then joins $p$ raking subgroup scans,
  allowing the subgroup-block scan to remain barrier-free and in registers. The subgroup-block reduction is then posted to shared memory. For a workgroup of size $w$, subgroup-raking reduces the size of the local spine to $w/s$.
  \item \textbf{Workgroup-Wide Scan}: The workgroup collectively performs a scan across the subgroup partial reductions, writing the results back to shared memory upon completion.
  \item \textbf{Chained Propagation, Decoupled Fallback}: The first subgroup in the workgroup performs a lookback operation to acquire the reduction result of all previous workgroups, engaging in fallback operations as necessary. Once completed, this subgroup posts the results to shared memory for the workgroup to consume.
  \item \textbf{Write Phase}: Each subgroup incorporates the results from the workgroup-wide scan and lookback operation to produce and write out the correct output.
\end{enumerate}
Phase 0 ensures that the work tiles are consumed in serial order. Phases 1, 2, 3, and 5 comprise the intra-workgroup scan, following a \emph{Scan-then-Propagate} strategy. Phase 4, \emph{Decoupled-Fallback}, is responsible for managing all inter-workgroup dependencies. As our loading and writing phase are typical of scan kernels, we focus here on our portability-specific adaptations.

@@

\subsection{Workgroup-Wide Scan}
The WebGPU specification supports subgroup sizes $s$, where $s = 2^k$ and $k \in [2, 7]$; on some hardware, subgroup sizes can vary between kernel launches. To handle this variability, we use a generalized radix-$s$ Ladner-Fischer~\cite{} scan network embedded with Kogge-Stone subgroup scans to perform the scan across the local spine. Because the upsweep phase of the Ladner-Fischer network follows a Brent-Kung~\cite{1675982} construction, assuming shared memory bank width equal to the subgroup size, any subgroup scans beyond the first will experience maximum $s$-way bank conflicts. To mitigate this, we employ a Merrill-Grimshaw~\cite[Section 3.3.5]{Merrill2009} style conflict avoidance strategy, which reduces bank conflicts to the radix base of the network—in our case, $s$. Counterintuitively, the resulting $s$-way conflict differs from the conflicts incurred by the Brent-Kung construction. Rather than multiple threads accessing different indices in the same memory bank, this conflict arises from a fanout originating from a single index. However, such access is interpreted by the hardware as a \emph{broadcast} operation, resulting in no conflicts. Although the Merrill-Grimshaw avoidance doubles our shared memory requirements, this increase causes no performance degradation due to our modest overall use of shared memory, such as our deliberate retention of vectors in registers and the efficiency of subgroup raking. Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes.\footnote{Since we also require a workgroup-wide pure reduce operation, we obtain this by omitting the fanout operation and substituting the subgroup Kogge-Stone scan with a subgroup butterfly reduction.}

\subsection{Decoupled-Fallback}
To reiterate, the goal of \emph{Chained Propagation} is twofold: first, to determine the reduction of all preceding work tiles, and second, to relay those values to succeeding tiles to accelerate their own lookback operations. Following the \emph{Decoupled-Lookback} technique, we model the inter-workgroup spine as a finite-state machine, where each work tile is represented by a reduction value and a state, stored in global memory. Each tile may exist in one of three states:
\begin{itemize}
  \item \textbf{Not Ready}: The tile has not yet been processed or written to global memory.
  \item \textbf{Ready}: The tile has completed its local reduction and posted it to global memory.
  \item \textbf{Inclusive}: The tile has completed its lookback and updated its posting to reflect the inclusive reduction.
\end{itemize}
Workgroups operate independently; each is responsible for updating its own tile state but may also opportunistically update preceding tiles in the event of a fallback. Tile states transition in a strictly forward direction---Not Ready $\rightarrow$ Ready $\rightarrow$ Inclusive---and never revert. Before kernel launch, all tiles are initialized to \emph{Not Ready} with reduction values set to $\epsilon$. State transitions imply an update of the reduction value: a transition to \emph{Ready} results in the posting of the local workgroup reduction, and to \emph{Inclusive}, the reduction of all previous workgroup reductions. To coordinate this process across the workgroup, each workgroup initializes a shared memory \emph{control flag} during partition index acquisition. This flag governs both the initiation of fallback operations and the entry and exit of the entire \emph{Decoupled-Fallback} phase, which begins immediately after the workgroup-wide scan over partial reductions is completed.

@@

  \item \textbf{Initialization:}
        \begin{itemize}
          \item The control flag, initialized to \emph{Locked}, signals that the workgroup should enter the \emph{Chained-Propagation} phase.
          \item The first thread initializes a \emph{previous reduction} variable in registers with $\epsilon$ to hold the accumulated reductions of preceding tiles.
          \item The first thread initializes a \emph{current tile} pointer to its immediate predecessor, with which it begins to traverse backwards through the spine.
        \end{itemize}

  \item \textbf{Lookback:}
        \begin{itemize}
          \item The first thread queries the state of \emph{current tile}.
                \begin{itemize}
                  \item If the tile is in the \emph{Ready} state, the workgroup ingests the value and sets \emph{current tile} to the preceding tile, repeating this step until reaching a tile that is \emph{Not Ready}.
                  \item If the tile is in the \emph{Inclusive} state, the workgroup ingests the value and proceeds directly to step~4.
                  \item If the tile is in the \emph{Not Ready} state, the first thread spins for a limited number of iterations, defined by the \emph{Maximum Spin Count}.
                        \begin{itemize}
                            \item If this count is exceeded, it exits the lookback loop and proceeds to step~3, leaving the control flag \emph{Locked} to signal to the workgroup that a fallback is required. Additionally, it copies \emph{current tile} into shared memory to broadcast the fallback tile identity to the workgroup.
                            \item Otherwise, it continues spinning until the tile transitions to a \emph{Ready} or \emph{Inclusive} state. Without FPG, \emph{Decoupled Lookback} risks spinning indefinitely and deadlocking at this stage.
                        \end{itemize}
                \end{itemize}
        \end{itemize}

  \item \textbf{Fallback:}
        \begin{itemize}
          \item If the blocking tile is still \emph{Not Ready} after waiting, the workgroup initiates a fallback operation.
          \item Using the tile index provided by the first thread, the full workgroup computes the reduction of the blocking tile by directly processing the corresponding elements. This process involves a \emph{workgroup-wide reduction},\footnote{See footnote NUMBER.} utilizing additional shared memory to aggregate partial subgroup results.
          \item Once the fallback reduction is complete, the workgroup \emph{attempts} to post the reduction to global memory and update the tile's state to \emph{Ready}.
          \item As this update attempt is performed atomically, the workgroup simultaneously reads back the blocking tile's most recent state:
                \begin{itemize}
                  \item If that most recent state is \emph{Inclusive}, the fallback value computed by the workgroup is discarded, and the first thread ingests the returned value before proceeding to step 4.
                  \item Otherwise, the first thread ingests the fallback value.
                        \begin{itemize}
                            \item If the blocking tile is the first in the spine, the workgroup has reached the start of the spine and has thus completed the lookback. The first thread proceeds to step 4.
                            \item Otherwise, the first thread updates \emph{current tile} to the preceding tile, returning to the lookback process in step~2.
                        \end{itemize}
                \end{itemize}
        \end{itemize}
